Point transect distance sampling in inlabru

### 2019-08-14 paper.tex
1st draft of introduction complete

Introduction outline:
\begin{enumerate}
	\item Estimation of abundance and spatial distribution of animals is an important topic
	\item Distance sampling is a family of methods that aim to do this whilst accounting for imperfect detectability of animals
	\item Distance sampling began life under a design-based inference paradigm
	\item Randomised survey design is used in two ways.  Firstly, to justify the assumption that the true distribution of animals is uniform with respect to distance from the observer.  Secondly, to construct Horvitz-Thompson-like estimators of animal density, which are then converted to abundance estimates for regions of interest
	\item Interest in the spatial distribution of animals has led to the development of approaches that combine traditional distance sampling methods and spatially explicit generalized additive models
	\item In this approach the estimated detection probabilities for discrete spatial units are used as an offset in a generalized additive model where the response variable is a count of detections within each unit
	\item This two-stage approach needs a way to propagate uncertainty from the detection model to the spatial model
	\item Recently [[JOYCE PAPER]] presented a one-stage approach that simultaneously estimates the detection and spatial processes, avoids binning data into counts and removes the need to consider uncertainty propagation
	\item This approach considered line-transect distance sampling as thinned log-Gaussian Cox process
	\item To estimate the detection function [[JOYCE et al]] constructed a spline-like function using an SPDE approach and fit the model using \texttt{R-INLA}
	\item In this paper we show how to achieve a similar one-stage model for point transect distance sampling
	\item Point transect surveys require a different approach than line transects because the area surveyed increases with distance from observer
	\item In addition to accounting for this [RJC - what does 'this' refer to?] we present an alternative way to specify the detection function.  We replace the spline approach with a parametric family of detection functions such as half-normal and hazard rate functions [RJC - I thought we couldn't use the hazard rate (hr) function? If the updated inlabru package includes the hr detection function model I need to rerun dsm as hr better fits the data.]
	\item  These detection functions are typically not linear in their parameters and so cannot be straightforwardly estimated using methods for additive linear predictors
	\item To solve this we implement an iterative approach based on a Taylor approximation of the detection function and fit the model using the \texttt{inlabru} package which extends the functionality of \texttt{R-INLA}
	\item This represents a novel development in distance sampling methods that allows simultaneous estimation of parametric detection functions and the spatial distribution of animals from point transect data
\end{enumerate}

@@

Over time interest grew in fully model-based approaches which allow the use of non-randomized survey designs by considering a spatial model for animal density so observations can be associated with spatially indexed covariates and abundance can, in principle, be estimated for any subregion within the study area \citep{johnson_model-based_2010, miller_spatial_2013, buckland_model-based_2016}.  Model-based distance sampling has typically been implemented in a two-stage modelling process.  In the first stage, detectability is estimated within each sampling unit.  In the second stage, detections within sampling units are binned into counts and used as the response variable in a generalized additive model framework where the detectability estimates from the first stage are used as an offset in the second stage.  Due to often sparse nature of wildlife survey data this may require consideration of over or under-dispersion and zero-inflated distributions.  Negative-binomial and Tweedie distributions are common choices to model the distribution of counts \citep{wood_gam_2017}.
A key concern with the two-stage approach is the propagation of the uncertainty in the detectability estimates to the second-stage spatial model.  Early attempts to address this focused on bootstrapping \citep{lahiri_resampling_2003}, but more recent work with the bootstrapping approach has pointed to potentially irresolvable difficulties that require resampling blocks at multiple spatial scales. Instead \cite{bravington_reliable_2018-1} proposed propagating error based on a second-order Taylor approximation of detectability around the maximum-likelihood estimate from the first stage.

\citet{yuan_point_2017} presented a one-stage approach for analysing line-transect distance sampling within a Bayesian framework that simulateously estimates the detectability 
and spatial distribution of animals.  Key to their approach is the use of sparse Gaussian Markov random field (GMRF) approximations to continously indexed Gaussian random fields (GRF) and a log-Gaussian Cox point process model that avoids the need to bin data into counts and naturally sidesteps issues of zero-inflation and dispersion.  The observation process is formulated as a thinning of the log-Gaussian Cox process.  Both the log-intensity of points and the detection function were modelled using a stochastic partial differential equation (SPDE) approach \citep{lindgren_explicit_2011}.  The model was specified as a latent Gaussian model and fitted using integrated nested Laplace approximations using the \texttt{R-INLA} package \citep{rue_approximate_2009}.

Here we present a similar approach to \citet{yuan_point_2017} but instead focus on the case of point transect data.  Point transects require special treatment to account for the fact that the area surveyed increases with increasing distance from the observer.  Our case study is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.  However, the point process viewpoint is not new and has been applied numerous times to line transect data [[citations here]].  We also present the first use of sparse GMRF with point transect distance sampling data.

We demonstrate an application of a new iterative model fitting approach that allows the inclusion of model components that are non-linear in their parameters.  This allows us to model detectability with a parametric family of detection functions, an approach that more closely resembles standard distance sampling methods than the SPDE formulation presented in \citet{yuan_point_2017}.

@@

\sloppy For the case with imperfect detection of points we specify a thinning probability function $g(s) = \mathbb{P}(\text{a point at $s$ is detected})$. A key property of the log-Gaussian Cox process is that a realisation thinned with thinning probability function $g(s)$ also follows a log-Gaussian Cox process with intensity given by $\tl(s) = \lambda(s)g(s)$.

@@

TODO HERE:  Description of Akepa data, survey method, key questions, what do we care about from ecology/conservation point of view.  Then, code snippets of how to specify the model in inlabru, brief discussion of PC priors and how they were chosen here

@@

TODO HERE:  Something about how this is all a one-stage model and all these plots are based on sampling from the joint posterior of all parameters.  As such all plots can be viewed as incorporating the uncertainty from the detection model.  This is a lovely and natural way to present results and an advantage of having a joint model.  I'm not sure varprop can do this? [RJC - Varprop does do this through the posterior covariance $V_beta$ matrix of the intercept, smoother and detection probability terms. But it doesn't do so in such an elegant way. Varprop is still reliant on a two-stage model.]

>>>>

@@

Over time interest grew in fully model-based approaches which allow the use of non-randomized survey designs by considering a spatial model for animal density.  This allows observations to be associated with spatially indexed covariates and abundance can, in principle, be estimated for any subregion within the study area \citep{johnson_model-based_2010, miller_spatial_2013, buckland_model-based_2016}.  
Model-based distance sampling has typically been implemented in a two-stage modelling process.  In the first stage, detectability is estimated within each sampling unit.  In the second stage, detections within sampling units are binned into counts and used as the response variable in a generalized additive model framework where the detectability estimates from the first stage are used as an offset.  Due to often sparse nature of wildlife survey data this may require consideration of over or under-dispersion and zero-inflated distributions.  Negative-binomial and Tweedie distributions are common choices to model the distribution of counts \citep{wood_gam_2017}.

A key concern with the two-stage approach is the propagation of the uncertainty in the detectability estimates to the second-stage spatial model.  Early attempts to address this focused on bootstrapping \citep{lahiri_resampling_2003}, but more recent work with the bootstrapping approach has pointed to potentially irresolvable difficulties that require resampling blocks at multiple spatial scales. Instead \cite{bravington_reliable_2018-1} proposed propagating error based on a second-order Taylor approximation of detectability around the maximum-likelihood estimate from the first stage.

\citet{yuan_point_2017} presented a one-stage approach for analysing line-transect distance sampling within a Bayesian framework that simulateously estimates the detectability 
and spatial distribution of animals.  Key to their approach is the use of a sparse Gaussian Markov random field (GMRF) approximation to a continously-indexed Gaussian random field (GRF) and a log-Gaussian Cox point process model that avoids the need to bin data into counts and naturally sidesteps issues of zero-inflation and dispersion.  The observation process is formulated as a thinning of the log-Gaussian Cox process.  Both the log-intensity of points and the detection function were modelled using a stochastic partial differential equation (SPDE) approach \citep{lindgren_explicit_2011}.  The model was specified as a latent Gaussian model and fitted using integrated nested Laplace approximations using the \texttt{R-INLA} package \citep{rue_approximate_2009}.

Here we present a similar approach to \citet{yuan_point_2017} but instead focus on the case of point transect data and replace the SPDE specification of the detection function with a parametric family of detection functions which are familiar to users of standard distance sampling methods.  Point transects require special treatment to account for the fact that the area surveyed increases with increasing distance from the observer.  Our case study is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.  However, the point process viewpoint is not new and has been applied numerous times to line transect data \citep{buckland_model-based_2016, johnson_model-based_2010, hedley_spatial_2004, stoyan_remark_1982, hogmander_random_1991}.  We also present the first use of a sparse GMRF spatial random effect with point transect distance sampling data.  The sparsity of the GMRF is useful for computational efficiency and the finite element mesh motivated by the SPDE allows efficient handling of complex spatial domains such as coastlines \citep{simpson_going_2016}.

To demonstrate the benefits of this one-stage approach we highlight potential issues regarding model evaluation and communication of results that are common in many formulations of species distribution modelling.  We discuss potentially misleading summary maps of the posterior intensity process and propose an alternative based on excursion sets \citep{bolin_excursion_2015, bolin_calculating_2018}.  A key theme in our case-study is to highlight the importance of considering \textit{realizations} of random quantities rather than rely too much as summaries such as expectations and quantiles.  The one-stage approach is invaluable since model summaries are based on monte-carlo sampling from the joint posterior of all model parameters.  Since the spatial process and observation process are estimated jointly this means figures and summaries produced naturally average over the estimated uncertainty in the observation processes.  This is an advantage over the two-stage approach even if uncertainty has been effectively propagated in some way.

The article proceeds as follows.  First we present the perspective of point transect distance sampling as a thinned poisson process, describing in detail how this can be formulated as a latent Gaussian model with a latent GMRF random effect.  We also show how the point process viewpoint can still be used in the case of incomplete location information where only the distance from the observer is recorded, not the exact location of the animal.  Secondly we describe how to specify the joint model using an iterative approach to fitting models with components that are non-linear in their parameters.  Thirdly, we introduce our case-study using a point transect distance sampling survey of the Hawaiian \akepa{} and show how the above methods can be specified and fit using the software package \texttt{inlabru} \citep{bachl_inlabru_2019}, an extension of the \texttt{R-INLA} package. We focus our model evaluation on the effective understanding and communication of uncertainty in the posterior intensity and abundance estimates, which are typically of greatest interest to stakeholders.

@@

\sloppy For the case with imperfect detection of points we specify a thinning probability function $g(s) = \mathbb{P}(\text{a point at $s$ is detected})$. A key property of the log-Gaussian Cox process is that a realisation of a point process with intensity process $\lambda(s)$ that is thinned by thinning probability function $g(s)$ also follows a log-Gaussian Cox process with intensity given by $\tl(s) = \lambda(s)g(s)$.

@@

TODO HERE:  Add code snippets of how to specify the model in inlabru, brief discussion of PC priors and how they were chosen here

@@

TODO HERE:  Something about how this is all a one-stage model and all these plots are based on sampling from the joint posterior of all parameters.  As such all plots can be viewed as incorporating the uncertainty from the detection model.  This is a lovely and natural way to present results and an advantage of having a joint model.  I'm not sure varprop can do this? [RJC - Varprop does do this through the posterior covariance $V_{\beta}$ matrix of the intercept, smoother and detection probability terms. But it doesn't do so in such an elegant way. Varprop is still reliant on a two-stage model.]
